What We Eval in the Shadows

### 2021-08-09 paper/main.tex
Move Shapes Redux to end and add Code Obfuscation section

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are the tools given to developers to
this end. In this section, we focus on the \emph{why}, giving real-world
examples of the \eval usage in the wild. All the examples are based on code from
our corpus, often simplified to fit the space and increase clarity. For each
example, we indicate the package and the function where it is located.

@@

%\todo{Say somewhere that the R community commonly calls call/argument capturing
%(with substitute or match.call), Non-Standard Evaluation.}
To understand the how R developers use \eval, we manually inspected top 117 eval
call-sites which contribute to 90\% of the eval calls. We describe the common
usage patterns below.

\textit{Core function variants.} Many packages use \eval to implement variations of the
\code{base} R functions. For example, the \code{igraph} package provides the
\code{do_call} function shown below. This is a variant of the \code{do.call}
function from the \code{base} package. This function executes a function call by
constructs it from a name and a list of arguments and calling \eval.
\begin{lstlisting}
do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
  f <- substitute(f)
  call <- make_call(f, ..., .args)
  eval(call, .env)
}
\end{lstlisting}

The \code{glue} package provides functions for string interpolation; text enclosed
by c{\{} and c{\}} is evaluated using \eval as an R expression and replaced by
its result. The code snippet below shows an example of this.

@@

Finally, let us mention an anecdotal use of \eval as a means to bypass some
checks performed by the \code{R CMD CHECK} tool. This tool performs a number of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example,  it checks that packages
do not call the \code{unlockBinding} function.\footnote{It allows one to open and
modify loaded package namespaces.} Packages such as \code{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.

>>>>

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 \eval call-sites which contribute to 90\% of the
\eval calls. In this section, we give examples of the \eval usage from this
real-world sample. The examples are often simplified to fit the space and
increase clarity. For each example, we indicate the package and the function
where it is located.

\emph{Core function variants.} Many packages use \eval to implement variations of the
\code{base} R functions. For example, the \code{igraph} package provides the
\code{do_call} function shown below. This is a variant of the \code{do.call}
function from the \code{base} package. This function executes a function call by
constructs it from a name and a list of arguments and calling \eval.

\begin{lstlisting}
do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
  f <- substitute(f)
  call <- make_call(f, ..., .args)
  eval(call, .env)
}
\end{lstlisting}

\emph{Domain-Specific Languages.} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation make R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
this.

text enclosed by c{\{} and c{\}} is
evaluated using \eval as an R expression and replaced by its result.

\emph{Remove Boilerplate}

For example, the PopED::get_rse uses it to assign a number of variables in the current scope: eval(parse(text=paste(default_args[[i]]), "← ", i)). There are also more sophisticated cases. The plyr::allocate_column function uses eval for a deferred assignment to fill missing values in a column. Package overture repeatedly evaluates R expressions capturing any assignments that happen in it as samples of Markov chains.

\emph{Code Obfuscation} Some packages use \eval to bypass the restrictions
imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. If package authors cannot avoid using these functions, they use \eval
to obfuscate the code, bypassing the checks performed by the tool. An example of
this use is the \code{scalaEnsure} function from the \code{aibd} package. It
mutates the variable \code{s} in the package environment. However, the package
environment in R is locked by default, which prevents bindings from being
mutated. Environment can be unlocked using \code{unlockBinding}, but its use is
flagged by the well-formedness checks. To work around this limitation, the
author passes the expression to unlock the package environment as a string to
\code{parse} which is then evaluated by \code{eval}.

\begin{lstlisting}
scalaEnsure <- function() {
    ...
    env <- parent.env(environment())
    eval(parse(text=paste0('unlockBinding("s", env)')))
    assign("s", s, envir=env)
    lockBinding("s", env)
}
\end{lstlisting}

\emph{Miscellaneous}

@@

%\todo{Say somewhere that the R community commonly calls call/argument capturing
%(with substitute or match.call), Non-Standard Evaluation.}

@@
